---
title: "Lab Exercise 1"
author: "Priyanka Verma"
date: today
date-format: "DD/MM/YY"
format: 
    pdf:
      toc: true
---

<!-- ## Shiny Documents -->

<!-- This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>. -->

<!-- ## Inputs and Outputs -->

<!-- <!-- You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive: -->

<!-- ```{r} --> <!-- sliderInput("bins", "Number of bins:",  --> <!--             min = 1, max = 50, value = 30) --> <!-- plotOutput("distPlot") --> <!-- ``` -->

<!-- ```{r} -->

<!-- #| context: server -->

<!-- output$distPlot <- renderPlot({ -->

<!--    x <- faithful[, 2]  # Old Faithful Geyser data -->

<!--    bins <- seq(min(x), max(x), length.out = input$bins + 1) -->

<!--    hist(x, breaks = bins, col = 'darkgray', border = 'white', -->

<!--         xlab = 'Waiting time to next eruption (in mins)', -->

<!--         main = 'Histogram of waiting times') -->

<!-- }) -->

<!-- ``` -->

```{r}
#install.packages("tidyverse")
library(tidyverse)
#library(ggplot2)
```

### Read Data

```{r}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm)
```

### Q1

Plot the ratio of male to female mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme

```{r}
d_1 <- dm |> 
  filter(Age==30 | Age==40 | Age==10 | Age==20 ) |>
  mutate(mf_ratio = Male/Female) |> 
  select(Year, Age, mf_ratio)
d_1
```

```{r}
d_1 |> 
  ggplot(aes(x=Year, y=mf_ratio, color= Age)) +
  geom_line()+
  ylab("Ratio male to female mortality rate") +
  ggtitle("Ratio of male to female mortality rates over time")+
  theme_light() #change of theme 

```

### Q2

Find the age that has the highest female mortality rate each year

```{r}
d_2 <- dm |> 
  select(Year, Age, Female) |>
  group_by(Year) |>
  mutate(maxF = max(Female, na.rm = TRUE)) |>
  group_by(Year) |>
  filter(Female == maxF) |>
  select(Year, Age, Female) # prints the Age along with the value of the highest Female mortality rate for each year
d_2
```

### Q3

Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations.

```{r}
d_3 <- dm  |>
  group_by(Age) |> 
  summarize(across(Female:Total, sd))
d_3
```

### Q4

The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here.

```{r}
dp <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 2, col_types = "dcddd")
head(dp)
```

```{r}
df_join <- merge(x = dm, y = dp, by = c("Year", "Age"), all.x = TRUE)
colnames(df_join) <- c("Year","Age","Female_m", "Male_m", "Total_m","Female_p","Male_p","Total_p")

df_join <- df_join |> 
  select(-c("Total_m", "Total_p")) # remove the total mortality and total population columns  
```

```{r}
df_join <- df_join |> 
  group_by(Year) |>
  mutate(total_f = sum(Female_p, na.rm = TRUE)) |>
  mutate(total_m = sum(Male_p, na.rm = TRUE))
```

```{r}
df_join <- df_join |>
  mutate(w_f= Female_m * Female_p/total_f) |>
  mutate(w_m = Male_m * Male_p/total_m)
```

```{r}
df_avg <- df_join |>
  group_by(Year) |>
  summarise(female = sum(w_f, na.rm = TRUE), male = sum(w_m, na.rm = TRUE))
```

```{r}
d_plot <- df_avg |> 
  pivot_longer(female:male, names_to = "Sex", values_to = "weighted_avg")
```

```{r}
p1 <- d_plot |> 
  ggplot(aes(x=Year, y=weighted_avg, color= Sex)) +
  geom_line()+
  ggtitle("Population weighted average mortality rate for males and females")+
  ylab("Population weighted average mortality rate") + 
  theme_light() #change of theme 
p1
```
Interpretation: 
The population weighted average mortality rate decreases and then increases for both females and males. 